Aggregations in R:
----

* There are many functions that can do aggregations for you in R; we will cover `ddply()` from the `plyr` package in this tutorial. This is also the function that we have found most useful when writing aggregate indicators for NMIS.

### Creating simple aggregated summary:
* note: 
 1. __(group) by__ variable must have at least one input
 2. You __must__ specify what type of aggregation you want to perform, choose one from: summarize, transform, your own complicated function 

 3. [The link to the package dodument](http://cran.r-project.org/web/packages/plyr/plyr.pdf)



```{r}
sample_data <- read.csv("~/work/r/nigeria_r_training/sample_health_facilities.csv")
library(plyr)
```

* First example, we use the "transform" keyword to specify adding aggregation to the right of original data
```{r}
my_summary <- ddply(sample_data, .(state, lga), transform, 
                    counts = length(lga_id),
                    total_num_nurse = sum(num_nurses_fulltime, na.rm=T),
                    avg_c_section = mean(c_section_yn == T,na.rm=T))
head(my_summary)
```

* Look at the output and compare the difference, the only change here is replacing summarize with transform
```{r}
my_summary <- ddply(sample_data, .(state, lga), summarise, 
                    counts = length(lga_id),
                    total_num_nurse = sum(num_nurses_fulltime, na.rm=T),
                    avg_c_section = mean(c_section_yn == T,na.rm=T))
head(my_summary)
```

* Ddply could take by variable in string format which is very handy when you want to use it in a function
```{r}
my_summary <- ddply(sample_data, c("state", "lga"), summarise, 
                    counts = length(lga_id),
                    total_num_nurse = sum(num_nurses_fulltime, na.rm=T),
                    avg_c_section = mean(c_section_yn == T,na.rm=T))
head(my_summary)
```

* Ddply could also take formulars wihch conforms the R convention
```{r}
my_summary <- ddply(sample_data, ~ state + lga, summarise, 
                    counts = length(lga_id),
                    total_num_nurse = sum(num_nurses_fulltime, na.rm=T),
                    avg_c_section = mean(c_section_yn == T,na.rm=T))
head(my_summary)
```


### Advanced ddply aggregation, with user defined functions:
* You are allowed/ encouraged to define your own function in ddply
* The syntax is pretty much the same as defining functions in R, except 
* It is __CRITICAL__ to add data.frame() function since ddply combines __data.frame__s, so make sure __include__ __data.frame()__ in the function definition
* for each small chunk of data splitted by ddply()

```{r}
my_summary <- ddply(sample_data, .(state), function(df){
              data.frame(
                unique_lga_number = nrow(df),
                avg_c_section = mean(df$c_section_yn == T,na.rm=T),
                avg_c_section_true = length(which(df$c_section_yn))
                        )})

head(my_summary)
```
### Advanced ddply aggregation, with user idata.frame functions:
* idata.frame comes from __plyr__ package, insted of making copies in memory, idata.frame simply creates a reference to the original object, thus to boost up the speed.
* If you have HUGE amount of data for ddply to process and you find it annoying to wait a long time before seeing the result, idata.frame is your solution.
* Please be noted it comes with the price of slightly complicated code.
* "An immutable data frame works like an ordinary data frame, except that when you subset it, it returns a reference to the original data frame, not a a copy. This makes subsetting substantially faster and has a big impact when you are working
with large datasets with many groups."

```{r}
isample <- idata.frame(sample_data)
my_summary <- ddply(isample, .(state), function(df){
                      data.frame(
                        unique_lga_number = nrow(df),
                        avg_c_section = mean(df$c_section_yn == T,na.rm=T),
                        avg_c_section_true = length(which(df$c_section_yn))
                      )
                    })
head(my_summary)
```


* Check the time, the difference with grow bigger with BIG dataset.
* Note: the first example doesn't actually work, we'll cover it in the next topic.
```{r, warning=FALSE}
system.time(replicate(100, ddply(isample, .(state), summarise, mean(num_nurses_fulltime, na.rm=T))))

system.time(replicate(100, ddply(sample_data, .(state), summarise, mean(num_nurses_fulltime, na.rm=T))))
```

* Draw backs of idata.frame: sometimes certain functions doesn't work with idata.frame
```{r}
my_summary <- ddply(isample, .(state), summarise, length(which(c_section_yn)))

my_summary <- ddply(sample_data, .(state), summarise, length(which(c_section_yn)))
```

### Excercise: How would you calculate the proportion of of c_section_yn==TRUE versus total non-NA records in each state?
* since the length(which()) doesn't work, what might be the candidate function go get counts of instances of TRUE and FALSE
* hint from day1: we learned table(), nrow()
* use table() to get frequency count of TRUE and FALSE values
* check if 'TRUE' is contained in the table() output, and assign 'TRUE' counts to numerator. If not return 0
* lastly getting the length of the c_section_yn column, do remember to use na.omit() to skim off the NA values.

```{r}
my_summary <- ddply(isample, .(state), function(df){ 
                    my_count <- table(df$c_section_yn)
                    data.frame(
                    num_true = if('TRUE' %in% names(my_count)){
                                my_count[['TRUE']]
                                }else{
                                    0
                                },
                    total_non_na = length(na.omit(df$c_section_yn))
                    )})
head(my_summary)
```

### Handy functions we created to do calculations with idata.frame

* icount(), function for counting the __TRUE__s in logical variable.
* Does it look familiar to the solution above?
* In order to get number of __FALSE__ simple evalute the logical variable.
```{r}
icount <- function(predicate) { 
    counts <- table(predicate)
    if('TRUE' %in% names(counts)) { counts['TRUE'] }
    else { 0 }
}

my_summary <- ddply(isample, .(state), function(df){ 
                    data.frame(
                    num_true_c_section = icount(df$c_section_yn),
                    num_true_c_section = icount(df$c_section_yn == FALSE), 
                    total_instances = nrow(df)
                    )})
head(my_summary)
```

* ratio(), function for getting ratio of two __logical__ varibles with filter.
* First argument is the numerator, second argument is the denominator, and the third argument is the __boolean filter_.
```{r}
ratio <- function(numerator_col, denominator_col, filter) {
    df <- data.frame(cbind(num=numerator_col, den=denominator_col))
    df <- na.omit(df[filter,])
    if (nrow(df) == 0 | sum(df$den) == 0){
        return(NA)
    }
    return(sum(df$num) / sum(df$den))
}


my_summary <- ddply(isample, .(state), function(df){ 
                    data.frame(
                    tru_false_ratio_public = ratio(df$c_section_yn == FALSE, 
                                                   df$num_nurses_fulltime >= 1, 
                                                   df$management == "public")
                    )})
head(my_summary)
```





