---
title: "Wind Analysis Bonus Simulation"
author: "Elliot Cohen"
date: "October 3, 2014"
output: html_document
---

********
## Bonus
### Stochastic Ensemble Forecast
To model windspeed for a given location, we fit a Weibull distribution to the observed data and then re-sample the fitted distribution to generate new realizations with the same distirubtional properties as the observed data.  Because every observation is a random realization from some underlying process, then it follows that any prediction based on the fitted distribution is similarly a random realization of the same underlying process.  Thus model predictions are probabilistic, not deterministic, and should be subject to probabalistic treatment.  Thefore, to "predict" next year's capacity factor, we will generate 500 simulations of probabilistic outcomes to better capture the range of likely outcomes.
```{r ensemble_forecast, cache=TRUE, eval=FALSE, include=FALSE}
windspeeds<-array(data=0, dim=c(500, 8760*6)) # blank array of winspeeds
gen.sim.mpc<-array(data=0, dim=c(500, 8760*6)) # blank array of energy output given windspeeds and the manufacturer's power curve (mpc)
gen.sim.epc<-array(data=0, dim=c(500, 8760*6)) # blank array of energy output given windspeeds and the empirical power curve (epc)

# populate the arrays with 500 simulated time series.
for (i in 1:500) {
  # Use a Weibull distribution to randomly generate one-year of 10-minute average windspeeds using the same shape and scale factor as the observed windspeeds
  windspeeds[i, ]<-rweibull(8760*6, shape=weibull.fit.stat.filter$estimate[1], scale=weibull.fit.stat.filter$estimate[2])
  
  # predict generation given new windspeeds using two different power curves
  gen.sim.mpc[i, ]<-predict(manufacturer.mod, newdata=windspeeds[i, ])
  gen.sim.epc[i, ]<-predict(empirical.mod, newdata=windspeeds[i, ])
  }    # end for-loop.

# organize and label simulation output
# initially, rows represent simulations (500), columns represent timesteps (10min)
# transpose such that rows are time, and columns are simulations
gen.sim.mpc<-t(gen.sim.mpc)
gen.sim.epc<-t(gen.sim.epc)

gen.sim.mpc<-as.data.frame(gen.sim.mpc)
gen.sim.epc<-as.data.frame(gen.sim.epc)

# create time series vector in ten-minute increments from 2013-01-01 to 2013-12-31
time<-seq(ISOdate(2013,1,1), to=ISOdate(2014,01,01), by=60*10)
gen.sim.mpc$POSIXct<-time[1:dim(gen.sim.mpc)[1]]
gen.sim.epc$POSIXct<-time[1:dim(gen.sim.epc)[1]]
```

Using the POSIX object, we can easily group the data with respect to time (e.g. by month, day, hour). This will come in handy for temporal aggregation, and for plotting:
```{r cut.2, eval=FALSE, include=FALSE}
month <- cut(gen.sim.mpc$POSIXct, breaks = "month")
week <- cut(gen.sim.mpc$POSIXct, breaks = "week")
day <- cut(gen.sim.mpc$POSIXct, breaks = "day")
hour <- cut(gen.sim.mpc$POSIXct, breaks = "hour")

dummy<-strsplit(as.character(week), split=" ")
week<-laply(dummy, '[[', 1) # keep the date, drop the time

dummy<-strsplit(as.character(day), split=" ")
day<-laply(dummy, '[[', 1) # keep the date, drop the time

dummy<-strsplit(as.character(hour), split=" ")
hour<-laply(dummy, '[[', 2) # keep the time, drop the date

gen.sim.mpc$hour<-as.factor(hour)
gen.sim.epc$hour<-as.factor(hour)
gen.sim.mpc$day<-as.factor(day)
gen.sim.epc$day<-as.factor(day)
gen.sim.mpc$week<-as.factor(week)
gen.sim.epc$week<-as.factor(week)
gen.sim.mpc$month<-as.factor(month)
gen.sim.epc$month<-as.factor(month)
```

```{r reshape, eval=FALSE, include=FALSE}
daily1<-ddply(gen.sim.mpc, .(day), numcolwise(sum))
daily2<-ddply(gen.sim.epc, .(day), numcolwise(sum))

# melt
dmelt1<-melt(daily1, id.vars="day")
dmelt2<-melt(daily2, id.vars="day")
dmelt<-merge(dmelt1, dmelt2, by="day")
```

```{r energy.timeseries.2, fig.width=9, fig.height=5, eval=FALSE, include=FALSE}
# # Now we can create time-series visualizations of the data:
# # summarize
# daily.gen.sim.mpc<-ddply(gen.sim.mpc, .(day), numcolwise(sum))
# dat<-melt(daily.gen.sim.mpc, id.vars="day")
# 
# # plot energy vs time
# test<-melt(energy, id.vars=("day"))
# ggplot(test, aes(x=day, y=value/10^3, group=variable, colour=variable, linetype=variable)) +
#   geom_line() +
#   scale_y_continuous(name="MWh per day") + 
#   labs(title="Energy Timeseries") +
#   theme_classic() +
#   scale_x_discrete(breaks=test$day[seq(1, 360, by=60)], labels=abbreviate)
# 
# # facet wrap by month
# energy<-subset(gen.sim.mpc, select=c("day","week", "month", "energy_sentout_10min_kwh", "wind_energy_10min_kwh", "betz_limit_10min_kwh", "uncurtailed_10min_kwh", "curtailment_10min_kwh"))
# energy<-ddply(energy, .(day, week, month), numcolwise(sum))
# 
# # plot energy vs time
# test<-melt(energy, id.vars=c("day", "week", "month"))
# levels(test$month) <- month.name[1:12]
# 
# ggplot(test, aes(x=day, y=value/10^3, group=variable, colour=variable, linetype=variable)) +
#   geom_line() +
#   facet_wrap(~month, scales="free") +
#   scale_y_continuous(name="MWh per day") + 
#   labs(title="Monthwise Energy Timeseries") +
#   theme_classic() +
#   scale_x_discrete(breaks=NULL)
```

```{r, eval=FALSE, include=FALSE}
# # # boxplot the hourly observations
# # colnames(gen.mpc)<-seq(ISOdate(2012,1,1), to=ISOdate(2013,1,1), by=60*10)
# # test<-t(gen.mpc) # hours are rows, columns are simulations
# # colnames(gen.mpc)<-"hours"
# # boxplot(test ~ hours)
# # 
# compute annual generation from the hourly estimated generation.
ann.gen.mpc<-apply(gen.sim.mpc, 1, sum)
ann.gen.epc<-apply(gen.sim.epc, 1, sum)
ann.gen<-cbind(ann.gen.mpc, ann.gen.epc)

# compute capacity factor in each year
cf<-ann.gen/(850*7*8760)

# show the range of estimates for capacity factor given the ensemble forecast
boxplot(cf, main="Capacity Factor Simulation", sub="Winspeed randomly sampled from a Weibull Distribution\nPower estimated from Manufacturer and Empirical Power Curves", ylab="Capacity Factor", ylim=c(0,1), names=c("manufacturer", "empirical"))
```